Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2024-10-30 src/tex/ms.tex
[showyourwork] overleaf sync

To infer $H_0$ from \ac{GW} sources, one needs not only the luminosity distance which can be inferred from the signal but also the redshift of the source.
Some \ac{GW} sources can be associated with \ac{EM} counterparts, allowing for the measurement of the redshift, e.g., GW170817 \citep{LIGOScientific:2017adf, Guidorzi:2017ogy}.
However, most \ac{GW} sources do not have \ac{EM} counterparts.
In this case, one can assume \ac{GW} sources are in galaxies and associate the redshift of the galaxy to the \ac{GW} source \citep{Schutz:1986gp, DelPozzo:2011vcw, Gray:2019ksv, Gray:2023wgj}.
Another alternative method is to use population statistics to infer $H_0$ by marginalizing the redshift distribution of the sources.
This method is known as the spectral-siren method \citep{Farr:2019twy, You:2020wju, Mastrogiovanni:2021wsd, LIGOScientific:2021aug, Ezquiaga:2022zkx}.

@@

Other studies have made use of non-parametric methods for population studies \citep[e.g.,][]{Mandel:2018mve, Tiwari:2020vym, Rinaldi:2021bhm, Edelman:2021zkw, Sadiq:2021fin, Edelman:2022ydv, Callister:2023tgi, Ray:2023upk}, and now these methods are starting to be applied to the spectral siren method.
For example, \cite{Farah:2024xub} used a Gaussian process to model the intrinsic mass distribution, which allows for a flexible model that can capture the features of the intrinsic distribution.

@@

The analysis code we developed to implement the method presented in this section is publicly available on GitHub\footnote{ \url{https://github.com/thomasckng/det-frame-cosmo-with-FIGARO/tree/main/src/scripts}}.

@@

In this section, we apply our method to the \ac{GW} events released by the \ac{LVK} as part of the GWTC-2.1 \cite{LIGOScientific:2021usb} and GWTC-3 \cite{KAGRA:2021vkt} catalogs.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}, these correspond to all the events with $\mathrm{FAR_{min}} < 1\, \mathrm{yr}^{-1}$, where $\mathrm{FAR_{min}}$ is the smallest \ac{FAR} among all search pipelines.
These events are different from the events used in \cite{LIGOScientific:2021aug}, where the authors use a \ac{FAR} threshold of $\mathrm{FAR_{min}} < 0.25\, \mathrm{yr}^{-1}$, therefore the results we present here might be slightly different from the ones in \cite{LIGOScientific:2021aug}.
We opted for a larger \ac{FAR} threshold -- consistent, however, with the one used in population studies such as \cite{KAGRA:2021duu} -- to have a larger number of events to include in the non-parametric analysis.

>>>>

To infer $H_0$ from \ac{GW} sources, one needs not only the luminosity distance which can be inferred from the signal but also the redshift of the source.
Some \ac{GW} sources can be associated with \ac{EM} counterparts, allowing for the measurement of the redshift, e.g., GW170817 \citep{LIGOScientific:2017adf, Guidorzi:2017ogy}.
However, most \ac{GW} sources do not have \ac{EM} counterparts.
In this case, one can assume \ac{GW} sources are in galaxies and associate the redshift of the galaxy to the \ac{GW} source \citep{Schutz:1986gp, DelPozzo:2011vcw, Gray:2019ksv, Mukherjee:2020hyn, Mukherjee:2022afz, Gray:2023wgj}.
Another alternative method is to use population statistics to infer $H_0$ by marginalizing the redshift distribution of the sources.
This method is known as the spectral-siren method \citep{Farr:2019twy, You:2020wju, Mastrogiovanni:2021wsd, LIGOScientific:2021aug, Ezquiaga:2022zkx, Karathanasis:2022rtr}.

@@

Other studies have made use of non-parametric methods for population studies \citep[e.g.,][]{Mandel:2016prl, Mandel:2018mve, Tiwari:2020vym, Rinaldi:2021bhm, Edelman:2021zkw, Sadiq:2021fin, Edelman:2022ydv, Callister:2023tgi, Ray:2023upk}, and now these methods are starting to be applied to the spectral siren method.
For example, \cite{Farah:2024xub} and \cite{MaganaHernandez:2024uty} used different non-parametric models to model the intrinsic mass distribution, which allows for a flexible model that can capture the features of the intrinsic distribution.

@@

The analysis code we developed to implement the method presented in this section and the results of the following sections are publicly available on GitHub\footnote{\url{https://github.com/thomasckng/det-frame-cosmo-with-FIGARO/tree/main/src/scripts}} and Zenodo \citep{ng_2024_13968239} respectively.

@@

In this section, we apply our method to the \ac{GW} events released by the \ac{LVK} as part of the GWTC-2.1 \cite{LIGOScientific:2021usb, ligo_scientific_collaboration_and_virgo_2022_6513631} and GWTC-3 \cite{KAGRA:2021vkt, ligo_scientific_collaboration_and_virgo_2023_8177023} catalogs.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}, these correspond to all the events with $\mathrm{FAR_{min}} < 1\, \mathrm{yr}^{-1}$, where $\mathrm{FAR_{min}}$ is the smallest \ac{FAR} among all search pipelines.
These events are different from the events used in \cite{LIGOScientific:2021aug}, where the authors use a \ac{FAR} threshold of $\mathrm{FAR_{min}} < 0.25\, \mathrm{yr}^{-1}$, therefore the results we present here might be slightly different from the ones in \cite{LIGOScientific:2021aug}.
We opted for a larger \ac{FAR} threshold -- consistent, however, with the one used in population studies such as \cite{KAGRA:2021duu} -- to have a larger number of events to include in the non-parametric analysis.
